---
title: "HMS 520 Final Project"
author: "Cassidy Chang, Yaz Ozten, and Sinclair Carr"
output: pdf_document

---

```{r}
rm(list = ls())
library("data.table")
library("ggplot2")


```




### Part I: Data Wrangling
  - Change names to be more readable and intuitive
  - CD4 (upper and lower) - multiply by 10
  - Sum CD4 variables into one to get average CD4 count
  - Separate age variable (lower limit and upper limit)
  - Create a variable indicating the length of study

```{r}

data <- setDT(read_excel("on-art_mort.xlsx"))
setnames(data, c("nid", "super", "meas_value", "meas_stdev", "iso3", "time_point", 
                 "age", "CD4_lower", "CD4_upper"), c("study_id", "gbd_superregion", 
                "p_mortality", "sd_p_mortality", "country_code", "time_since_art", 
                "age_group", "cd4_lower", "cd4_upper"))

data[, `:=` ("cd4_lower" = cd4_lower * 10, "cd4_upper" = cd4_upper * 10)]
data[, study_length := time_upper - time_lower]
data[, cd4_mid := (cd4_upper + cd4_lower)/2]
data[, sex := as.factor(ifelse(sex == 1, "male", "female"))]

# Identify missing values 

colnames(data)[colSums(is.na(data)) > 0]
sum(is.na(data$p_mortality)) # 20
sum(is.na(data$sd_p_mortality)) # 20
sum(is.na(data$site)) # 360

View(data)

```

  
### Part II: Data Analysis
  - Descriptive analysis
  - Attempt k-means/agglomerative/spectral clustering and see if predictive of time since ART 
  - Running different regression models with different sets of covariates and mortality and use of ART as outcomes
  - Comparing models with select goodness of fit tests

```{r}
model1 <- p_mortality ~ sex + age + gbd_superregion
```


### Part III: Visualization
  - Plot regression outputs 
  - Plot goodness of fit tests
  - Visualization with PCA and TSNE 

```{r}




```


