---
title: "HMS 520 Final Project"
author: "Cassidy Chang, Yaz Ozten, and Sinclair Carr"
output: pdf_document

---

```{r}
rm(list = ls())
library("data.table")
library("ggplot2")


```




### Part I: Data Wrangling
  - Change names to be more readable and intuitive
  - CD4 (upper and lower) - multiply by 10
  - Sum CD4 variables into one to get average CD4 count
  - Separate age variable (lower limit and upper limit)
  - Create a variable indicating the length of study

```{r}

data <- setDT(read_excel("on-art_mort.xlsx"))
setnames(data, c("nid", "super", "meas_value", "meas_stdev", "iso3", "time_point", 
                 "age", "CD4_lower", "CD4_upper"), c("study_id", "gbd_superregion", 
                "p_mortality", "sd_p_mortality", "country_code", "time_since_art", 
                "age_group", "cd4_lower", "cd4_upper"))

data[, `:=` ("cd4_lower" = cd4_lower * 10, "cd4_upper" = cd4_upper * 10)]
data[, study_length := time_upper - time_lower]
data[, cd4_mid := (cd4_upper + cd4_lower)/2]
data[, sex := as.factor(ifelse(sex == 1, "male", "female"))]

# Identify missing values 

colnames(data)[colSums(is.na(data)) > 0]
sum(is.na(data$p_mortality)) # 20
sum(is.na(data$sd_p_mortality)) # 20
sum(is.na(data$site)) # 360

View(data)

```

  
### Part II: Data Analysis
  - Descriptive analysis
  - Attempt k-means/agglomerative/spectral clustering and see if predictive of time since ART 
  - Running different regression models with different sets of covariates and mortality and use of ART as outcomes
  - Comparing models with select goodness of fit tests

```{r}
summary(data, digits = 2)
by(data, data$gbd_superregion, summary)
by(data, data$age_group, summary)
table(data$sex, data$time_since_art)
prop.table(table(data$sex, data$time_since_art)) # Summaries of general descriptive statistics based on GBD super region and age group, as well as proportions of participants in the data set based on time since ART.

boxplot(data$p_mortality~data$age_group) # Age and probability of mortality may be slightly correlated
boxplot(data$p_mortality~data$time_since_art) # Much of the data seems to be clustered in the range of six months since ART. Probability of mortality seems to decrease as time since ART increases (from six to 12 to 24 months)

model2 <- lm(formula = p_mortality ~ sex + age_group + time_since_art, data = data)
library("rigr")
regress("odds", formula = p_mortality ~ sex + age_group + time_since_art, data = data) # The above seems to be more or less true as we can see by running the linear model of probability of mortality against sex, age group and time since ART

model1 <- p_mortality ~ sex + age + gbd_superregion
```


### Part III: Visualization
  - Plot regression outputs 
  - Plot goodness of fit tests
  - Visualization with PCA and TSNE 

```{r}




```


